---
title: "Calculating Annualized Carbon Estimates"
author: "Eric Scott"
format: gfm
toc: true
---

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(tidyr)
library(here)

```

I'm going to try to work through a small sample of data to figure out how the code in forestTIME-builder works

# Read in data

I'll use RI as an example.
Some of the columns get auto-detected as the wrong type (e.g. CN should be character, not numeric, even though it looks like a number).
I got these column types from the Green Book and recorded them in a list in `scripts/create_column_types.R`.

```{r}
tree_raw <- read_csv(
  here("data/rawdat/state/RI_TREE.csv"),
  col_types = cols(
    CN = col_character(),
    PLT_CN = col_character(),
    PREV_TRE_CN = col_character(),
    DAMTYP1 = col_integer(),
    DAMSEV1 = col_integer(),
    DAMTYP2 = col_integer(),
    DAMSEV2 = col_integer(),
    ACTUALHT = col_double(),
    HT = col_double(),
    DIA = col_double(),
    MORTYR = col_double(),
    INVYR = col_double()
  )
)

plot_raw <- read_csv(
  here("data/rawdat/state/RI_PLOT.csv"),
  col_types = cols(
    CN = col_character()
  )
)

plotgeom_raw <- read_csv(
  here("data/rawdat/state/RI_PLOTGEOM.csv"),
  col_types = cols(
    CN = col_character()
  )
)

#needed later
cond_raw <- read_csv(
  here("data/rawdat/state/RI_COND.csv"),
  col_types = cols(
    CN = col_character(),
    PLT_CN = col_character(),
    NBR_LIVE_STEMS = col_integer(),
    LAND_COVER_CLASS_CD_RET = col_integer(),
    TRTYR2 = col_integer(),
    DSTRBYR2 = col_integer()
  )
)
```

# Initial Wrangling

## Tree table

Tree table gets filtered to only year 2000 and onward, the `CN` column is renamed to be more specific, and composite IDs are created that should be universally unique (not just unique within the plot or state)

```{r}
tree <- tree_raw |> 
  filter(INVYR >= 2000.0) |> 
  rename(TREE_CN = CN) |> 
  mutate(
    PLOT_COMPOSITE_ID = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"),
    TREE_COMPOSITE_ID = paste(STATECD, UNITCD, COUNTYCD, PLOT, SUBP, TREE, sep = "_"),
    .before = 1
  )

```

It is my understanding that the `TREE_COMPOSITE_ID` is now consistent through time as well since the combination of state, unit, county, plot, sub-plot, and tree number are unique for each tree and the same every inventory.
But if that's try, what does the `chain_by_joins()` function do??

### Chain by joins

::: callout-important
## Question

It looks like `chain_by_joins()` adds a `TREE_FIRST_CN` column that connects every observation to the first `CN` a tree was given.
Is this column needed?
Or is `TREE_COMPOSITE_ID` enough?

Answer: I found references in Renata's notes to two methods for creating unique keys for trees and that they work equally well for *most* trees but sometimes disagree.
I wonder if sometimes trees at the border of plots "move" plots between years or something like that?
:::

## Plot table

I need to join one column from the PLOTGEOM table (`ECOSUBCD`) that is needed for the carbon estimation code (it used to be in the PLOT table in older versions of FIADB)

We also do similar filtering and renaming as in the tree table

```{r}
#only care about keys and the ECOSUBCD column
plotgeom <- plotgeom_raw |> select(CN, INVYR, ECOSUBCD)
plot <- left_join(plot_raw,
                  plotgeom,
                  by = join_by(CN, INVYR)) |>
  filter(INVYR >= 2000.0) |>
  rename(PLT_CN = CN) |>
  mutate(
    PLOT_COMPOSITE_ID = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"),
    .before = 1
  )
```

## Cond table

Needed later on for carbon estimation

```{r}
cond <-
  cond_raw |>
  filter(INVYR >= 2000.0) |>
  rename(COND_CN = CN) |>
  mutate(
    PLOT_COMPOSITE_ID = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"),
    .before = 1
  )
```

# Annualizing

I'm skipping some steps in `create_all_tables.R`, but I don't think they're all necessary.
This section would replace what goes on in `add_annual_estimates_to_db.R`, which I think is probably overly complicated and has the issue of not working when a tree goes "missing" for a survey.

First, `ACTUALHT` and `HT` get coalesced.
If I remember correctly, `ACTUALHT` only applies when a tree is broken, and we want to use that value for carbon estimation if it exists.

```{r}
tree <- 
  tree |> 
  mutate(ACTUALHT = coalesce(ACTUALHT, HT)) |> 
  select(ACTUALHT, HT, everything())
```

Maybe I'm not understanding something that the code needs to do, but I think we can just \~\~use the built-in R function for linear interpolation, `approx()`.

Let's try an example

### Toy Example

```{r}
#| label: toy-data
df <- tibble::tribble(
  ~TREE_COMPOSITE_ID, ~YEAR, ~HT, ~STATUSCD,
  1, 2000, 20, 1,
  1, 2005, 30, 1,
  1, 2010, 37, 1,
  1, 2015, 45, 2,
  2, 2000, 5, 1,
  2, 2005, NA, NA,
  2, 2010, NA, NA,
  2, 2015, 20, 1,
  2, 2020, 32, 1,
  3, 2000, NA, 2,
  4, 2000, 10, 1,
  4, 2005, 14, 1,
  4, 2010, NA, 2
)

```

There are two trees each measured every 5 years.
Tree 1 was measured in 4 surveys, the last of which it was dead.
Tree 2 was missing for two surveys, and then found again still alive.
Tree 3 was only recorded once and it was already dead in the first survey.

First we need to expand it to include all years between surveys.

```{r}
#| label: toy-expand
all_yrs <-
  df |>
  group_by(TREE_COMPOSITE_ID) |>
  expand(YEAR = full_seq(YEAR, 1))

df_all_yrs <- right_join(df, all_yrs) |> arrange(TREE_COMPOSITE_ID, YEAR)
df_all_yrs
```

Then, I think we just need to use `approx()` in a `mutate()` call, but let's make a "friendlier" version that returns a vector and does nothing when it can't interpolate (i.e. it needs at least 2 non-`NA` values to interpolate).

We also want linear *extrapolation* of trailing `NA`s in the case that a "mort tree" (tree first recorded dead) has no observations for `DIA` or `HT`.
That is accomplished with `Hmisc::approxExtrap()`, but needs a continuous set of numbers created by `approx()` to start with.

```{r}
#| label: interpolate-function
inter_extra_polate <- function(x, y) {
  if (sum(!is.na(y)) < 2) {
    return(y)
  } else {
    #first interpolate
    interpolated <- approx(x, y, xout = x)$y
  }
  #then extrapolate trailing NAs if needed
  if (all(!is.na(interpolated))) {
    return(interpolated)
  } else {
    extrapolated <- 
      Hmisc::approxExtrap(
        x = x[!is.na(interpolated)],
        y = interpolated[!is.na(interpolated)],
        xout = x[is.na(interpolated)]
      )$y
    return(c(interpolated[!is.na(interpolated)], extrapolated))
  }
}

#for "interpolating" categorical vars
step_interp <- function(x) {
  if (all(is.na(x))) {
    return(x)
  }
  x_non_nas <- x[!is.na(x)]
  i <- seq_along(x)
  
  # index at which each non-missing value is
  i_non_nas <- i[!is.na(x)]
  
  # count leading NAs
  leading_NAs <- i_non_nas[1] - 1
  
  #this is the number of rep()s on the right side (not including the non-missing value)
  right <- floor((dplyr::lead(i_non_nas) - i_non_nas) / 2)
  right[is.na(right)] <- 0
  
  #number of rep()s to the left, including the non-missing value
  left <- ceiling((i_non_nas - dplyr::lag(i_non_nas)) / 2)
  left[is.na(left)] <- 1
  
  times <- left + right
  
  #extrapolate trailing NAs
  #reps of last value gets adjusted so sum(times) adds up to the total length of the input
  times[length(times)] <- length(x) - (sum(times) - times[length(times)] + leading_NAs)
  
  x_interp <- purrr::map2(x_non_nas, times, rep) |> purrr::list_c()
  
  #add back leading NAs
  c(rep(NA, leading_NAs), x_interp)
}
```

```{r}
#| label: toy-interpolate
df_interpolated <- df_all_yrs |> 
  group_by(TREE_COMPOSITE_ID) |> 
  mutate(HT_interpolated = inter_extra_polate(x = YEAR, y = HT))
df_interpolated
```

Yeah, that seems to have done it!
Then we can just cut off any dead trees at the estimated mortality year.

```{r}
#| label: toy-mortyr
df_interpolated <- 
  df_interpolated |> 
  group_by(TREE_COMPOSITE_ID) |> #for each tree...
  #figure out when it was last recorded alive and first recorded dead
  mutate(
    last_live = YEAR[max(which(STATUSCD == 1))],
    first_dead = YEAR[min(which(STATUSCD == 2))]
  ) |> 
  #estimate the mortality year as the midpoint between surveys
  mutate(
    dead_yr = ceiling(mean(c(last_live, first_dead))),
    last_yr = if_else(!is.na(dead_yr), dead_yr, max(YEAR))
  ) |> 
  #remove estimates after death
  filter(YEAR <= last_yr) |> 
  select(-last_live, -first_dead, -dead_yr, -last_yr)
```

```{r}
#| label: toy-plot
library(ggplot2)
ggplot(df_interpolated, aes(x = YEAR)) +
  geom_point(aes(y = HT)) +
  geom_line(aes(y = HT_interpolated)) +
  facet_wrap(vars(TREE_COMPOSITE_ID))
```

### Actual data

It *might* be trickier to do with the real data because:

1.  We might need to "annualize" the `TREE_CN` key, which is different for each inventory year for the same `TREE_COMPOSITE_ID`
2.  Some, but not many, trees have a recorded `MORTYR`, which we'd like to (optionally) use instead of the midpoint between surveys as an estimated end to that tree's timeseries.

Renata's code only interpolates/extrapolates `DIA`, `HT`, and `ACTUALHT` and other variables needed for carbon estimation are added later with joins.
I'm not sure that's the best approach, especially since some of the other variables used (e.g. `CULL`, the precentage of rotten wood) are continuous and could also be linearly interpolated.
I'm going to try doing the necessary joins *first* and then annualizing.

Variable used by carbon estimation code in `add_carbon_variables_*.R`:

Interpolation needed:

-   `CR` - `tree` table; compacted crown ratio (interpolate)
-   `DIA` - `tree` table; diameter (interpolate)
-   `HT` - `tree` table; height (interpolate)
-   `ACTUALHT` - `tree` table; height accounting for broken tops (interpolate)
-   `CULL` - `tree` table; rotten and missing cull, a percentage (interpolate)
-   `STATUSCD` - `tree` table; alive (1) or dead (2)
-   `DECAYCD` - `tree` table; needed for joining to `REF_TREE_DECAY_PROP` (categorical). `NA` for live trees, 1-5 for dead trees
-   `STANDING_DEAD_CD` - `tree` table; `NA` for live trees, 0 or 1 for dead trees.
-   `STDORGCD` - `cond` table; stand origin code, method of stand regeneration for trees in the condition (2.5.25) (categorical). This is also `NA` sometimes, but here I think it is safe to assume that `NA` can be replaced with `0` (natural stands).
-   `COND_STATUS_CD` - `cond` table; eventually for carbon estimation only trees in 'accessible forest land' (`COND_STATUS_CD == 1`) are kept (not sure why)

Invariant; join *after* interpolation:

-   `CULL_DECAY_RATIO` - this is a weird one—it's the `DENSITY_PROP` value from the `REF_TREE_DECAY_PROP` table when `DECAYCD = 3`. I do not know why, but it's in the David Walker code like this.
-   `ECOSUBCD` - `plot` table; ecological subsection code (invariant)
-   `SPCD` - `REF_SPECIES` table; species code (invariant[^1])
-   `SFTWD_HRDWD` - `REF_SPECIES` table; softwood or hardwood (invariant)
-   `JENKINS_SPGRPCD` - `REF_SPECIES` table; species group code (invariant)
-   `WDSG` - renamed from `WOOD_SPGR_GREENVOL_DRYWT` in `REF_SPECIES` table; green specific gravity of wood based on volume attributes in the `tree` table (invariant)
-   `CARBON_RATIO_LIVE` - `REF_SPECIES` table; wood carbon fraction (invariant)
-   `CARBON_RATIO` - `REF_TREE_CARBON_RATIO_DEAD` table; wood carbon fraction of dead trees (invariant)


```{r}
tree |> 
  group_by(TREE_COMPOSITE_ID) |> 
  filter(length(unique(STANDING_DEAD_CD)) > 1) |> 
  select(TREE_COMPOSITE_ID, INVYR, STATUSCD, STANDING_DEAD_CD, everything()) |> 
  arrange(TREE_COMPOSITE_ID, INVYR)
```


There are some trees in the `tree` table that were dead on their first survey and don't have any height or diameter measurements.
We should get rid of these.
I'll also drop trees that change species (for now)

```{r}
#| label: tree-filter-always-dead
tree <- tree |> 
  group_by(TREE_COMPOSITE_ID) |> 
  filter(!(all(is.na(DIA)) |
             all(is.na(ACTUALHT)) |
             all(is.na(HT)) |
             all(STATUSCD != 1)))

# drop trees with more than 1 SPCD
tree <- tree |> 
  group_by(TREE_COMPOSITE_ID) |> 
  filter(length(unique(SPCD)) == 1)
```

Let's convert `NA`s for `DECAYCD` and `STANDING_DEAD_CD` to `-1`s (an unused code) *for now*, so the variable works nicely with our `step_interp()` function (i.e. to distinguish `NA`s for live trees or missing inventories from years that need interpolation), then we can switch `-1`s back to `NA`s later based on estimated mortality year.

```{r}
tree <- tree |> 
  mutate(across(c(DECAYCD, STANDING_DEAD_CD), \(x) if_else(is.na(x), -1, x)))
```

And let's assume an `NA` for `STDORGCD` is a `0`

```{r}
cond_cols <- cond |> 
  select(PLOT_COMPOSITE_ID, INVYR, CONDID, COND_STATUS_CD, STDORGCD) |> 
  mutate(STDORGCD = if_else(is.na(STDORGCD), 0, STDORGCD))

tree <- 
  left_join(tree, cond_cols,
            by = join_by(PLOT_COMPOSITE_ID, INVYR, CONDID))
```

#### Interpolation & Extrapolation

First, interpolate/extrapolate continuous variables

```{r}
tree_to_interpolate <- tree |> 
  select(
    TREE_COMPOSITE_ID,
    PLOT_COMPOSITE_ID,
    SPCD,
    INVYR,
    DIA,
    HT,
    ACTUALHT,
    CR,
    CULL,
    STATUSCD,
    MORTYR,
    DECAYCD,
    STANDING_DEAD_CD,
    STDORGCD,
    COND_STATUS_CD
  )
```

First, expand to include all years

```{r}
#| label: tree-expand

# this needs to include all columns that are invariant over time
all_yrs <- tree_to_interpolate |> 
  group_by(TREE_COMPOSITE_ID, PLOT_COMPOSITE_ID, SPCD) |>
  expand(YEAR = full_seq(INVYR, 1))

tree_annual <- 
  right_join(
    tree_to_interpolate,
    all_yrs,
    by = join_by(TREE_COMPOSITE_ID, PLOT_COMPOSITE_ID, SPCD, INVYR == YEAR)
  ) |>
  arrange(TREE_COMPOSITE_ID, INVYR) |> 
  rename(YEAR = INVYR)
tree_annual
```

Then interpolate with our custom `approx()` wrapper function, `inter_extra_polate()`

```{r}
#| label: tree-interpolate
tree_interpolated <- 
  tree_annual |> 
  group_by(TREE_COMPOSITE_ID) |> 
  mutate(
    across(c(ACTUALHT, DIA, HT, CULL, CR),
           \(var) inter_extra_polate(x = YEAR, y = var)),
    across(c(DECAYCD, STANDING_DEAD_CD, STDORGCD, COND_STATUS_CD), step_interp)
  )
tree_interpolated
```

After interpolating, we can change those -1s back to NAs

```{r}
tree_interpolated <- tree_interpolated |> 
  mutate(across(c(DECAYCD, STANDING_DEAD_CD), \(x) if_else(x == -1, NA, x)))
```

There are some trees where interpolation fails.
I suspect they are all ones with only one non-NA measurement and there is nothing we can do.

```{r}
failed_to_interpolate <- 
  tree_interpolated |> 
  filter(any(is.na(DIA))) |>
  pull(TREE_COMPOSITE_ID) |>
  unique()
tree |> 
  group_by(TREE_COMPOSITE_ID) |> 
  filter(!all(STATUSCD != 1)) |> 
  filter(TREE_COMPOSITE_ID %in% failed_to_interpolate) |>
  arrange(TREE_COMPOSITE_ID, INVYR) |> 
  select(TREE_COMPOSITE_ID, INVYR, STATUSCD, DIA, ACTUALHT, everything())
```

#### `REF_SPECIES` table

Join in time-invariant columns that are required for carbon estimation

```{r}
# REF_SPECIES table variables
ref_species_raw <- read_csv(here("data/rawdat/REF_SPECIES.csv"))
ref_species <- ref_species_raw |> 
  select(SPCD,
         JENKINS_SPGRPCD, 
         SFTWD_HRDWD, 
         CARBON_RATIO_LIVE,
         WDSG = WOOD_SPGR_GREENVOL_DRYWT)
tree_interpolated <- left_join(tree_interpolated, ref_species, by = join_by(SPCD))
```

#### `ECOSUBCD` from the `plot` table

```{r}
plot_ecosubcds <- plot |>
  select(PLOT_COMPOSITE_ID, ECOSUBCD) |>
  distinct()

tree_interpolated <- left_join(tree_interpolated, plot_ecosubcds, by = join_by(PLOT_COMPOSITE_ID))
```

#### REF_TREE_DECAY_PROP

::: callout-important
## Question

I'm not sure what's going on here, but it *looks* like the code creates a new column called `CULL_DECAY_RATIO` using the `DENSITY_PROP` column for `DECAYCD` 3.
Then later it looks like this is overwritten if the tree is NOT alive (`STATUSCD!=1`) to be 1.
There are no notes in the code (either Renata's or David Walker's) indicating why this is done.
`DECAYCD` 3:

> Only limb stubs exist; the top is broken; a variable amount of bark remains; sapwood is sloughing; heartwood has advanced decay in upper bole and is beginning at the base.
:::

```{r}
ref_tree_decay_prop <- read_csv(here("data/rawdat/REF_TREE_DECAY_PROP.csv"))
ref_tree_carbon_ratio_dead <- read_csv(here("data/rawdat/REF_TREE_CARBON_RATIO_DEAD.csv"))
```

```{r}
tree_interpolated <- 
  tree_interpolated |> 
  ungroup() |> 
  #first joins by SFTWD_HRDWD only the DENSITY_PROP column for DECAYCD 3, but calls it CULL_DECAY_RATIO
  left_join(
    ref_tree_decay_prop |>
      filter(DECAYCD == 3) |>
      select(
        SFTWD_HRDWD,
        #not sure I understand the naming of this variable
        CULL_DECAY_RATIO = DENSITY_PROP
      ),
    by = join_by(SFTWD_HRDWD) 
  ) |> 
  #then joins additional columns (including DENSITY_PROP) based on DECAYCD and SFTWD_HRDWD
  left_join(
    ref_tree_decay_prop |>
      select(SFTWD_HRDWD,
             DECAYCD,
             DENSITY_PROP,
             BARK_LOSS_PROP,
             BRANCH_LOSS_PROP),
    by = join_by(DECAYCD, SFTWD_HRDWD)
  ) |> 
  #then join to the carbon ratio dead table to get CARBON_RATIO
  left_join(
    ref_tree_carbon_ratio_dead |>
      select(SFTWD_HRDWD, DECAYCD, CARBON_RATIO),
    by = join_by(DECAYCD, SFTWD_HRDWD)
  ) 
```

#### Additional wrangling

`NA`s for `CULL` get replaced with 0s (not entirely sure why, but assumes that no data = 100% live wood)

```{r}
tree_interpolated <- tree_interpolated |> 
  mutate(CULL = ifelse(is.na(CULL), 0, CULL))
```


### Visualize results

Let's take a sample of trees and visualize them.

```{r}
set.seed(123)

tree_sample <- tree_interpolated |> pull(TREE_COMPOSITE_ID) |> unique() |> sample(50)
tree_subset <- tree_interpolated |>
  filter(TREE_COMPOSITE_ID %in% tree_sample)
tree_subset |> 
  select(TREE_COMPOSITE_ID, YEAR, DIA, HT, ACTUALHT, CR, CULL, DECAYCD, STANDING_DEAD_CD, STDORGCD) |> 
  pivot_longer(DIA:STDORGCD, names_to = "variable") |> 
  ggplot(aes(x = YEAR, y = value, color = TREE_COMPOSITE_ID)) +
  facet_wrap(vars(variable), scales = "free_y") +
  geom_line(show.legend = FALSE)
```

This makes me think about caveats of interpolation since it's not really appropriate to linearly interpolate ACTUALHT—trees tops don't break off linearly.

# Mortality

We need to estimate mortality of trees in interpolated data using two methods:

1. Assume the trees died at the midpoint (rounding up) between the last year they were recorded alive and the first year they were recorded dead
2. Using the `MORTYR` variable if it was recorded, otherwise with #1 above

::: callout-important

I think this has to happen *before* carbon estimation because standing dead trees contribute carbon.  That means there does need to be two versions of the resulting annualized carbon estimates I think.

:::

#### Estimate mortality

Then we need to determine the year trees died.
For now, let's maybe store this in a different table that we can use in a filtering join later.

The estimated mortality year will be used for:

1. Determining when to switch `STATUSCD` from 1 (alive) to 2 (dead)\
2. Determining when to use `NA`s for `STANDING_DEAD_CD` (non-`NA`s only for dead trees)
2. Determining when to use `NA`s for `DECAYCD` (non-`NA`s only for standing dead trees >4.9 `DIA`)
3. Determining when to use `NA`s for `DENSITY_PROP`, `BARK_LOSS_PROP`, and `BRANCH_LOSS_PROP`, which were joined on `DECAYCD`

I think the strategy that will be most efficient is to produce a separate table with mortality estimations for each tree that can be used as input to a function that does some joining and mutating.

```{r}
#| label: tree-mortyr

#NOTE: some of these `if_else()` could be replaced with `coalesce()`, but `if_else()` might be more explicit and readable

tree_last_yr <-
  tree_interpolated |> 
  group_by(TREE_COMPOSITE_ID) |> #for each tree...
  #figure out when it was last recorded alive and first recorded dead
  mutate(
    last_live = YEAR[max(which(STATUSCD == 1))],
    first_dead = YEAR[min(which(STATUSCD == 2))]
  ) |> 
  #estimate the mortality year as the midpoint between surveys
  mutate(
    #TODO confirm that midpoint rule rounds up?
    dead_yr = ceiling(mean(c(last_live, first_dead))),
    #if dead_yr is NA still (because it was never alive and mean(c(NA, first_dead)) is NA), then just use the first_dead year
    dead_yr = if_else(is.na(dead_yr), first_dead, dead_yr),
  #use MORTYR data if it exists
    dead_yr_mortyr = if_else(!is.na(MORTYR), MORTYR, dead_yr),
    death_yr_midpt = if_else(!is.na(dead_yr), dead_yr, max(YEAR)),
    death_yr_mortyr = if_else(!is.na(dead_yr), dead_yr_mortyr, max(YEAR))
  ) |> 
  #summarize per tree
  group_by(TREE_COMPOSITE_ID) |> 
  summarize(death_yr_midpt = max(death_yr_midpt, na.rm = TRUE),
            death_yr_mortyr = max(death_yr_mortyr, na.rm = TRUE))

```

#### Adjust for mortality

```{r}

#test tree
# tree_example <- tree_interpolated |> 
#   filter(TREE_COMPOSITE_ID == "44_1_1_228_3_7")

#join in the mortality year estimates
tree_interpolated_midpoint <- left_join(
  tree_interpolated, tree_last_yr,
  by = join_by(TREE_COMPOSITE_ID)
) |>
  select(TREE_COMPOSITE_ID,
         YEAR,
         death_yr_midpt,
         STATUSCD,
         DECAYCD,
         everything()) |>
  #use midpoint rule column
  #if the tree isn't dead, set STATUSCD to 1, else to 2
  mutate(
    STATUSCD = if_else(YEAR >= death_yr_midpt, 2, 1),
    #if tree is alive, set STANDING_DEAD_CD to NA
    STANDING_DEAD_CD = if_else(YEAR < death_yr_midpt, NA, STANDING_DEAD_CD),
    #only use DECAYCD for standing dead trees >4.9 DIA
    DECAYCD = if_else(STANDING_DEAD_CD == 1 & DIA > 4.9, DECAYCD, NA)
  ) |> 
  # Set things like BRANCH_LOSS_PROP to NA as they were joined on DECAYCD
  mutate(across(
    c(DENSITY_PROP, BARK_LOSS_PROP, BRANCH_LOSS_PROP, CARBON_RATIO),
    \(var) if_else(is.na(DECAYCD), NA, var)
  )) |> 
  # clean up temporary vars
  select(-death_yr_midpt, -MORTYR)

#Some tests:
tree_interpolated_midpoint |> 
  filter(STATUSCD == 1 & !is.na(STANDING_DEAD_CD)) |>
  select(TREE_COMPOSITE_ID, YEAR, STATUSCD, STANDING_DEAD_CD, DECAYCD) |> 
  arrange(TREE_COMPOSITE_ID, YEAR)

tree_interpolated_midpoint |> 
  filter(STANDING_DEAD_CD == 0 & !is.na(DECAYCD)) |>
  select(TREE_COMPOSITE_ID, YEAR, STATUSCD, STANDING_DEAD_CD, DECAYCD) |> 
  arrange(TREE_COMPOSITE_ID, YEAR)
```



# Carbon estimation

Here's where we use the functions in `carbon_code/` to estimate carbon and AGB.

I'll be trying to replicate the code in `R/add_carbon_variables_mortyr.R`


Create and re-name variables needed for carbon estimation

::: callout-note
I think some of the odd looking choices here have to do with limitations in the carbon estimation code provided by David Walker.  Index-based filtering doesn't work when there are `NA`s.

E.g.

```{r}
#| error: true
df <- data.frame(x = c(1, 2, NA), y = 1:3)
df[df$x < 2]
```
:::

```{r}
tree_prepped <-
  tree_interpolated_midpoint |>
  mutate(
    #use 1 for CULL_DECAY_RATIO for live trees
    CULL_DECAY_RATIO = if_else(STATUSCD == 1, CULL_DECAY_RATIO, 1),
    # Using 0 instead of NA for live trees likely has to do with index-based
    # filtering
    STANDING_DEAD_CD = if_else(STATUSCD == 1, 0, STANDING_DEAD_CD),
    DECAYCD = if_else(STATUSCD == 1, 0, DECAYCD),
    # these are the proportion *remaining* after deducting for decay, so set to
    # 1 for live trees
    DECAY_WD = if_else(STATUSCD == 1, 1, DENSITY_PROP),
    DECAY_BK = if_else(STATUSCD == 1, 1, BARK_LOSS_PROP),
    DECAY_BR = if_else(STATUSCD == 1, 1, BRANCH_LOSS_PROP),
    #TODO: why is this called C_FRAC if it is a percentage?
    C_FRAC = if_else(STATUSCD == 1,
                     CARBON_RATIO_LIVE * 100,
                     CARBON_RATIO * 100)
  ) |> 
  #done with these columns
  select(-CARBON_RATIO_LIVE, CARBON_RATIO, DENSITY_PROP, BARK_LOSS_PROP, BRANCH_LOSS_PROP)
```

Filtering

::: callout-important
## Question

Why do we filter to exclude trees that are not alive or dead and standing?
Why only include `COND_STATUS_CD` 1 (accessible forest land)?
This removes 1,627 rows for RI
:::

```{r}
fiadb <- tree_prepped <- tree_prepped |> 
  #only keep accessible forest land (???)
  filter(COND_STATUS_CD == 1) |> 
  #remove fallen trees
  filter((STANDING_DEAD_CD == 1 & STATUSCD == 2) | STATUSCD == 1) |> 
  #remove trees with no recorded DIA (where interpolation failed I guess?)
  filter(!is.na(DIA)) |> 
  #done with COND_STATUS_CD column
  select(-COND_STATUS_CD)
```

With `fiadb` in the global environment, we can now run `add_carbon_variables_midpoint.R` successfully!! 🎉


#TODO test with `MORTYR`

Make sure all the codes get interpolated correctly when `MORTYR` is used as the first year a tree is dead.


[^1]: About 60 trees in RI do actually "change" species.
    These are likely misidentifications.
    More at <https://github.com/mekevans/forestTIME-builder/issues/53>
