---
title: "Un-interpolatable trees"
date: today
format: 
  html:
    toc: true
code-fold: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk: 
    out-width: 90%
    fig-width: 10
---

We currently drop trees from the dataset before interpolation for three main reasons:

1.  The tree changes `SPCD` at some point in it's life
2.  It was included by error---`RECONCILECD` is 7 or 8 at some point in it's life.  Trees that have `STATUSCD` 0 and an `NA` for `RECONCILECD` *might* also be included in this category.
3.  It only has one measurment for `HT` or `DIA` (even if it appears in multiple surveys)

1 and 2 are choices we've made (and could make differently), but 3 is due to a limitation—you can't "connect the dots" with interpolation if there's only one dot!

There are several (legit) reasons a tree might only have one measurement, each with different signals in the data.
We might approach these trees in different ways.

## Reasons a tree would only have one measurement

In rough order of "we know how to deal with this" to "🤷‍♂️"

1.  It was always fallen (I'm not sure if this is something that actually shows up in the data)
    -   Signal:
        -   Only `NA`s for `HT` and `DIA`
        -   `STANDING_DEAD_CD = 0`
    -   Solution: we would drop these anyway!
2.  It was alive or standing dead and then fell down before being measured again
    -   Signal:
        -   In one inventory, numbers for `HT` and `DIA`
        -   In the next inventory, `NA` for `HT` and `DIA`, `STATUSCD  = 2`, `STANDING_DEAD_CD = 0`
    -   Possible solution: Cary the last (only) measurements forward until interpolated death year
3.  The plot the tree is in has only been visited once
    -   Signal:
        -   The `plot_ID` and `tree_ID` should only show up in one year in the dataset
        -   `STATUSCD = 1` or `STANDING_DEAD_CD = 1` (alive or standing dead)
    -   Possible solutions:
        -   Only use base inensity plots (✅)
        -   Some trees may have a second measurement pre-2000---we could use the pre-2000 data for interpolation and filter it out after.
4.  Ingrowth
    -   Signal:
        -   The `tree_ID` only appears once, but `plot_ID` shows up more than once
        -   `DIA` < 10 or explicitly marked as ingrowth (`RECONCILECD` 1)
5.  Deforestation
    -   Signal:
        -   An observation with `COND_STATUS_CD` 2.
6.  Condition change (only one measrement in a sampled condition)
    -   Signal:
        -   More than one observation
        -   At least one `RECONCILECD` 9 
7.  Other (everthing left, there are sometimes a lot of trees that fit this category)
    -   Signal:
        - Only one measurement
        - Tree appears in only one inventory, but plot has been resampled
        - No indication of weirdness with `STATUSCD` 0 or a non-`NA` `RECONCILECD`.
        - Doesn't fit the definition of ingrowth above

## The data

I'll read in raw data for a few states, add our `tree_ID` and `plot_ID` composite ID columns, and select columns of interest.  I'll also filter to **only include base intensity plots**

```{r}
#| label: setup
#| include: false

library(forestTIME.builder)
library(tidyverse)
library(rFIA)

theme_set(theme_dark())

states <- c("OR", "CA", "AZ", "DE", "RI", "ID", "AK", "MA", "CT")
# states <-c("DE", "RI")

get_fia_tables(states = states, download_dir = here::here("fia"), keep_zip = TRUE)
```

```{r}
#| label: get-data
db <- readFIA(dir = here::here("fia"), states = states, tables = c("TREE", "COND", "PLOT"))
TREE <- db$TREE |>
  as_tibble() |>
  filter(INVYR >= 2000) |>
  add_composite_ids() |>
  left_join(state_codes, by = join_by(STATECD)) |>
  select(
    state_abb,
    plot_ID,
    tree_ID,
    CONDID,
    INVYR,
    HT,
    DIA,
    SPCD,
    STATUSCD,
    STANDING_DEAD_CD,
    RECONCILECD,
    CARBON_AG
  ) 

PLOT <-
    db$PLOT |>
    dplyr::filter(INVYR >= 2000L) |>
    add_composite_ids() |>
    dplyr::select(
      plot_ID,
      CN,
      INVYR,
      DESIGNCD, #for joining TPA_UNADJ based on rules later
      # MACRO_BREAKPOINT_DIA #unclear if this is really needed
      INTENSITY
    )

COND <-
    db$COND |>
    dplyr::filter(INVYR >= 2000L) |>
    add_composite_ids() |>
    dplyr::select(
      plot_ID,
      CONDID,
      INVYR,
      COND_STATUS_CD,
      COND_NONSAMPLE_REASN_CD
    )

tree <- left_join(TREE, COND) |>
  left_join(PLOT) |> 
  dplyr::filter(INTENSITY == 1) |> #only keep base intenstiy plots
  arrange(state_abb, tree_ID, INVYR)
```

I'll do my best to programatically assign trees to each of these categories

First, a quick question: are there trees with measurments for `HT` but not `DIA` or vice versa?

```{r}
bind_rows(
  tree |>
    filter(!is.na(HT) & is.na(DIA)),
  tree |>
    filter(!is.na(DIA) & is.na(HT))
)
```

Quite a few!
I'm working under the assumption that we need *both* `HT` and `DIA` to estimate carbon, so we would need to drop a tree that meets the "only one measurement" condition for *either* of these variables.

## Status Quo

Currently, this is how our "rules" exclude trees from the sample before interpolation & extrapolation

```{r}
#| label: functions

reasons_fct <-
  c(
    "Included by error" = "error",
    "Never observed standing" = "always_fallen",
    "Recorded species changes" = "spcd",
    "One or fewer measurements" = "only_one",
    "Plot hasn't been revisited" = "only_visit",
    "Fell after first inventory" = "fell",
    "Ingrowth" = "ingrowth",
    "Condition change" = "nonsampled",
    "Deforestation" = "deforestation",
    "Other" = "other"
  )

wrangle_dropped <- function(.data) {
  data_summary <- .data |>
    group_by(state_abb, INVYR, reason) |>
    summarize(
      n_dropped = length(unique(tree_ID)),
      carbon_dropped = sum(CARBON_AG, na.rm = TRUE),
    ) |>
    group_by(state_abb, INVYR) |>
    mutate(trees_total = sum(n_dropped), carbon_total = sum(carbon_dropped)) |>
    filter(!is.na(reason))

  reasons_fct <- reasons_fct[reasons_fct %in% unique(data_summary$reason)]

  data_summary |>
    ungroup() |> 
    mutate(reason = fct_relevel(reason, !!!unname(reasons_fct))) |> 
    mutate(reason_long = fct_recode(reason, !!!reasons_fct))
}


colors <- c(
  "One or fewer measurements" = "#88CCEE",
  "Never observed standing" = "#72190E",
  "Fell after first inventory" = "#DDCC77",
  "Plot hasn't been revisited" = "#117733",
  "Recorded species changes" = "#332288",
  "Included by error" = "#AA4499",
  "Ingrowth" = "#44AA99",
  "Condition change" = "#999933",
  "Deforestation" = "#882255",
  "Other" = "gray"
)

plot_dropped <- function(.data, y_var = c("n", "carbon")) {
  y_var <- match.arg(y_var)

  if (y_var == "n") {
    p <-
      ggplot(
        .data,
        aes(x = INVYR, y = n_dropped / trees_total, fill = reason_long)
      ) +
      labs(y = "Proportion of trees dropped")
  }

  if (y_var == "carbon") {
    p <-
      ggplot(
        .data,
        aes(x = INVYR, y = carbon_dropped / carbon_total, fill = reason_long)
      ) +
      labs(y = "Proportion of carbon in dropped trees")
  }
  p +
    facet_wrap(vars(state_abb)) +
    geom_col() +
    scale_fill_manual(values = colors) +
    scale_x_continuous(breaks = seq(2000, 2024, by = 2), minor_breaks = NULL) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.title = element_blank()
    )
}
```

```{r}
#| column: page
trees_dropped_current <- tree |>
  group_by(tree_ID) |>
  mutate(
    #case_when makes the reasons mutually-exclusive for plotting consistency
    #the conditions may not actually be mutually exclusive (pretty sure they aren't)
    #case_when evaluates in order, so conditions (try to be) most specific to most general
    reason = case_when(
      any(RECONCILECD %in% c(7, 8)) | all(STATUSCD == 0 & is.na(RECONCILECD)) ~ "error",
      !(sum(!is.na(DIA) & !is.na(HT)) >1) ~ "only_one",
      !(length(unique(SPCD)) == 1) ~ "spcd",
    )
  )

dropped_current <- wrangle_dropped(trees_dropped_current)

dropped_current |>
  plot_dropped(y_var = "n")
dropped_current |>
  plot_dropped(y_var = "carbon")
```

::: callout-warning
In some years, for some states, 100% of the trees would be excluded by our methods!
:::

::: callout-note
The totals for a given year (total trees, total carbon) are only for the plots visited in that year
:::

## Detailed breakdown

Let's see if we can break this down into the different reasons for having one or fewer measurements in the data.

::: callout-note
These categories are assigned in a mutually exclusive way---a tree can only belong to one of these categories.  I've taken care to assign them in an order that makes sense.  E.g. if a tree is recorded by error, it gets assigned that category regardless of whether it was also ingrowth or had a species change, because we would get rid of it anyways.
:::

```{r}
#| label: filter-dropped

# how many times does each plot_ID show up in the data?
plot_inv_count <-
  tree |>
  select(plot_ID, INVYR) |>
  distinct() |>
  count(plot_ID) |>
  rename(plot_inv_count = n)

# add logical columns for each reason for possibly being dropped
# NOTE: `TRUE` means dropped and *either* `FALSE` or `NA` means not dropped
trees_dropped_detailed <- tree |>
  left_join(plot_inv_count, by = "plot_ID") |>
  group_by(tree_ID) |>
  mutate(
    reason = case_when(
      #Earlier conditions take precedence over later ones
      #So, these are in order of "there's no solution, we will drop them anyways" then specific to general cases that have solutions.
      any(RECONCILECD %in% c(7, 8)) |
        all(STATUSCD == 0 & is.na(RECONCILECD)) ~ 
        "error",

      #fallen trees sometimes get diameter measurements, but I don't think ht is measured
      sum(!is.na(DIA) & !is.na(HT)) == 0 &
        all(STANDING_DEAD_CD == 0) &
        plot_inv_count > 1 ~
        "always_fallen",

      #tree fell over after being measured only once
      sum(!is.na(DIA) & !is.na(HT)) == 1 &
        n() > 1 &
        #treating dead with no measurements the same as fallen
        (any(STANDING_DEAD_CD == 0) | any(STATUSCD == 2 & is.na(DIA) & is.na(HT))) ~
        "fell",
      
      #tree was in a condition sampled in one inventory, but now in nonsampled condition
      #OR tree was in non-sampled condition and only once in sampled condtion
      sum(!is.na(DIA) & !is.na(HT)) <= 1 &
        n() > 1 &
        any(RECONCILECD == 9) ~
        "nonsampled",
      
      #plot and tree only exists once in dataset
      plot_inv_count == 1 & n() == 1 ~ "only_visit",

      #deforestation?
      sum(!is.na(DIA) & !is.na(HT)) <= 1 &
        # n() > 1 &
        any(COND_STATUS_CD == 2) ~ "deforestation",

      #Then, capture any remaining trees that have STATUSCD 0 with no explanation here
      #  sum(!is.na(DIA) & !is.na(HT)) <= 1 &
      #   any(STATUSCD == 0) &
      #   all(is.na(RECONCILECD)) ~ "no_status",
        

      #plot has been inventoried previously, small DIA or explicitly marked as ingrowth
      #this may also include trees that shrank (RECONCILECD = 5)
      drop_ingrowth = sum(!is.na(DIA) & !is.na(HT)) <= 1 &
        (any(DIA < 10) | any(RECONCILECD == 1)) &
          plot_inv_count > 1 ~
        "ingrowth",

      #changing species (we can fix this, so it's at the end)
      !(length(unique(SPCD)) == 1) ~ "spcd",

      #flag any remaining trees with only one observation and measurement
      sum(!is.na(DIA) & !is.na(HT)) == 1 &
      n() == 1 ~ "other"
    )
  )
```

Then we can summarize by state and inventory year to get a sense of how many trees are dropped, what proportion of total trees are represented by each category, and how much total carbon is represented by each category.


```{r}
dropped_detailed <- wrangle_dropped(trees_dropped_detailed)
```

```{r}
#| column: page
plot_dropped(dropped_detailed, y_var = "n")
```

```{r}
#| column: page
plot_dropped(dropped_detailed, y_var = "carbon")
```

There are a surprising number of trees that were only sampled once because of condition change (from sampled to non-sampled or from non-sampled to sampled).  Let's take a closer look at these

```{r}
trees_dropped_detailed |> 
  ungroup() |> 
  filter(reason == "nonsampled") |> count(COND_NONSAMPLE_REASN_CD)
```

- 1 = outside U.S. boundary
- 2 = denied access area
- 3 = hazardous situation
- 5 = lost data
- 6 = lost plot (can't find entire plot)
- 7 = wrong location (plot found, but it's not in the right spot?)
- 8 = skipped visit (entire plot skipped)
- 9 = dropped intensified plot
- 10 = other
- 11 = ocean
- `NA` = ???
