---
title: "Adding interpolated TPA_UNADJ"
author: Eric Scott
format: html
---

Experimenting with how we will add a `TPA_UNADJ` column to the annualized tree table.

```{r}
library(DBI)
library(duckdb)
library(dplyr)
library(tidyr)
library(readr)
library(here)
```

Setup connection and tables

```{r}
con <- dbConnect(duckdb(here("data/db/foresttime-from-state-parquet.duckdb")))
```

```{r}
(tables <- dbListTables(con))
tree <- tbl(con, "tree")
plot <- tbl(con, "plot")
tree_carbon <- tbl(con, "tree_carbon_annualized_midpoint")
```

What design codes actually have trees in them?

```{r}
left_join(tree, plot |> select(PLOT_COMPOSITE_ID, DESIGNCD)) |> 
  count(DESIGNCD) |> collect() |> knitr::kable()
```

Get TPA for design codes 558 and 559 which don't use the same TPA rules based on DIA, but are invariant through time.

```{r}
tree_designcd_tpa <- left_join(
  tree,
  plot |> select(PLOT_COMPOSITE_ID, DESIGNCD),
  by = join_by("PLOT_COMPOSITE_ID")
) |> select(TREE_COMPOSITE_ID, DESIGNCD, TPA_UNADJ) |> 
  #just keep one row per tree, ideally with a non-NA TPA_UNADJ
  group_by(TREE_COMPOSITE_ID, DESIGNCD) |> 
  summarize(TPA_UNADJ = max(TPA_UNADJ, na.rm = TRUE), .groups = "drop")

TPA_558_559 <- tree_designcd_tpa |> 
  filter(DESIGNCD %in% c(558, 559)) |> 
  select(TREE_COMPOSITE_ID, TPA_UNADJ)
```

Fill in TPA_UNDJ for design codes 558 and 559

```{r}
tree_carbon_tpa1 <- left_join(tree_carbon, TPA_558_559)

#checking that it worked
tree_carbon_tpa1 |> 
  filter(TREE_COMPOSITE_ID == "53_5_29_11_2_4871760") |> 
  select(TPA_UNADJ,)
```

Try out rules translated from Green Book by Dani in DESIGNCD_TPA.csv.
This uses a [rolling join](https://www.tidyverse.org/blog/2023/01/dplyr-1-1-0-joins/#rolling-joins).

```{r}
rules <- read_csv(here("data/DESIGNCD_TPA.csv")) |> 
  select(-...1)
#the rules table needs to be added to the database in order for joins to work. This can just be temporary though
copy_to(con, rules, "rules", temporary = TRUE, overwrite = TRUE)

tree_designcd <- tree_designcd_tpa |> select(-TPA_UNADJ)

#add the design code column
tree_carbon_tpa <- left_join(
  tree_carbon_tpa1,
  tree_designcd,
  by = join_by(TREE_COMPOSITE_ID)
) |> 
  #rolling join to add the rules
  left_join(tbl(con, "rules"), by = join_by(DESIGNCD, between(DIA, min_DIA, max_DIA))) |> 
  #fill in the TPA_UNADJ column if there isn't anything there
  mutate(TPA_UNADJ = if_else(is.na(TPA_UNADJ), TPA_2, TPA_UNADJ)) |> 
  select(-TPA_2)
  
#check that it worked (this took a lot of RAM with the arrange())
tree_carbon_tpa |> 
  select(TREE_COMPOSITE_ID, YEAR, TPA_UNADJ, everything()) #|> 
  # arrange(TREE_COMPOSITE_ID, YEAR)

```

It would probably be good to test this with a smaller dataset with known outcome.
