---
title: "Simulation"
author: Eric Scott
format: html
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
source(here::here("R/inter_extra_polate.R"))
source(here::here("R/step_interp.R"))
```

The idea here is to simulate what happens when you only periodically sample a population with a panel design and then interpolate using either the actual year of death of individuals or assume midpoint between samples.

## Simulated tree growth

I don't think it really matters much how I simulate tree growth for this, but two options are a logistic growth curve (plus observation error) or a Ricker model, which includes "process error" (i.e. different growth rates for each year) (plus observation error).

Logistic:

```{r}
# Generate a logistic growth curve with paramaters pulled randomly and with observation error
gen_logistic <- function(x = 1:100, obs_err = 2) {
  #not exactly sure what these parameters do, but fiddled around until I got mostly nice looking logistic growth curves.
  phi1 <- rnorm(1, 100, 5) #related to max height?
  phi2 <- rnorm(1, -10, 2) #related to starting height?
  phi3 <- rbeta(1, 1.2, 8) #related to growth rate?
  
  #logistic growth equation + random (normal) error
  y <- phi1 / (1 + exp(-(phi2 + phi3 * x))) + rnorm(x, sd = obs_err)
  y
}
```

```{r}
x <- 1:200
plot(x, gen_logistic(x, 2))
```

Ricker:

I pulled this equation from an [old blog post](https://ericrscott.com/posts/2020-02-24-hacking-glms/index.html#hacking-a-ricker-model-glm) of mine and it is probably not the most appropriate for trees, but I like the inclusion of "process error".

$$
N_{t+1} = N_te^{r(1-\frac{N_t}{K})}
$$

```{r}
# Generate logistic-ish growth with max height `K` and random growth rates each year, plus observation error
gen_ricker <- function(x = 1:100,
                       K = rnorm(1, 100, 20),
                       obs_err = 2) {
  rs <- rbeta(length(x), 1, 10) #random growth rates each year
  #repeatedly calculate Nt+1 using it for the next Nt with accumulate()
  Ns <- purrr::accumulate(
    rs,
    function(n_t, r) {
      n_t * exp(r * (1 - n_t / K)) 
    },
    .init = sample(1:10, 1)
  )
  # correct length of vector and add observation error
  Ns[seq_along(x)] + rnorm(length(x), 0, obs_err)
}
```

```{r}
x <- 1:200
plot(x, gen_ricker(x, obs_err = 2))
```

## Simulation setup

I created some parameters for number of trees per plot, number of plots, and then assign each tree a random time step for birth and random death time between 20 and 150 timesteps after birth

```{r}
n_trees <- 50
n_plots <- 20

sim_setup <-
  expand_grid(
    tree = seq_len(n_trees),
    plot = seq_len(n_plots)
  ) |> 
  #give each tree a unique ID and lifespan
  mutate(tree = paste0("p", plot, "t", tree)) |> 
  #give each tree a birth time and death time
  mutate(birth_time = sample(1:200, n(), replace = TRUE)) |> 
  mutate(death_time = birth_time + sample(20:150, n(), replace = TRUE))
sim_setup
```

## Mass mortality

Whether we use the *actual* death year or the midpoint between surveys to interpolate death should only really matter if there are known mass mortality events, so let's introduce one.
I'll pick 20% of living trees in a single plot in time step 73 (arbitrary, but chosen to *not* be a inventory year or midpoint—see below) and overwrite their death year.

```{r}
#get a random slice of 20% of trees in plot 1 that would be alive in timestep 73
mass_mortality_trees <- sim_setup |> 
  group_by(tree) |> 
  filter(plot == 1, between(73, birth_time, death_time)) |> 
  ungroup() |> 
  slice_sample(prop = 0.2) |> 
  pull(tree)

#overwrite the death time for those trees
sim_setup <- sim_setup |> 
  mutate(death_time = if_else(tree %in% mass_mortality_trees, 73, death_time))
```

## Run Simulation

Now I'll grow those trees using my Ricker model generator function (could use logistic too. Both are probably equally wrong)

```{r}
  #run simulation for all the years from the first birth time to last death time
sim <- sim_setup |> 
  group_by(tree, plot) |> 
  expand(time = birth_time:death_time) |>
  arrange(plot, tree, time) |> 
  #grow trees
  group_by(tree) |> 
  mutate(ht = gen_ricker(time)) |> 
  mutate(mortyr = max(time))
```

Let's see what it looks like:

```{r}
ggplot(sim, aes(x = time, y = ht, color = as.factor(plot), group = tree)) + 
  geom_point() +
  geom_line() +
  theme(legend.position = "none")
```

Let's only keep the middle years and scale them to start at 1 so births and deaths kinda even out??

```{r}
sim <- sim |> filter(time >= 100, time <= 200) |> 
  mutate(time = time - 99,
         mortyr = mortyr - 99)
```

```{r}
ggplot(sim, aes(x = time, y = ht, color = as.factor(plot), group = tree)) + 
  geom_point() +
  geom_line() +
  theme(legend.position = "none")
```

## Simulate Panel Design

Measure plots every 10 years, but in a way that the same number of plots are measured every year.

I'll just use the plot numbers as an offset

::: callout-note
Note that this means the choice of number of plots is related to this step.
If you change number of plots, this probably has to change too.
:::

Figuring out how to stagger the plot observations was the hardest part!

```{r}
plot = 1:20 #20 plots
period = 10 #measured every 10 years
length = 100 #for 100 years
seq(plot[1] %% period, length, period)
seq(plot[2] %% period, length, period)
#...
seq(plot[11] %% period, length, period) #same years as plot 1
seq(plot[12] %% period, length, period) #same years as plot 2

#create inventory schedule
inventory_schedule <- sim |> 
  #get all the years for all the trees
  select(plot, tree, mortyr) |>
  distinct() |>
  expand_grid(time = 1:101) |> 
  group_by(plot) |> 
  #filter to just inventory years
  filter(time %in% seq(unique(plot) %% period, 101, period)) |> 
  ungroup() |> 
  #keep the inventory after death
  filter(lag(time) < mortyr) |> 
  #expand again, this time for just the range of inventories for each particular tree
  group_by(tree, plot, mortyr) |> 
  expand(time = full_seq(time, 1)) |> 
  group_by(plot) |> 
  mutate(invyr = time %in% seq(unique(plot) %% period, 101, period))

periodic_expanded <- 
  #add extra years and invyr column
  right_join(sim, inventory_schedule) |> 
  arrange(plot, tree, time) |> 
  mutate(ht = if_else(invyr, ht, NA)) |> 
  #remove leading NAs
  group_by(tree) |> 
  filter(!all(is.na(ht))) |> 
  filter(row_number() >= which.min(is.na(ht)))


periodic <- periodic_expanded |> 
  #remove non-inventory years
  filter(invyr)
  
```

```{r}
ggplot(periodic, aes(x = time, y = ht, color = as.factor(plot), group = tree)) + 
  geom_point() +
  geom_line() +
  theme(legend.position = "none")
```

## Interpolation

### Use mortyr

This is an extreme case where we "know" the mortality year for every tree

```{r}
interpolated_mortyr <-
  periodic_expanded |> 
  group_by(plot, tree) |> 
  #linearly interpolate ht
  mutate(ht = inter_extra_polate(time, ht)) |> 
  #filter dead trees
  filter(time < mortyr)
```

```{r}
ggplot(interpolated_mortyr, aes(x = time, y = ht, color = as.factor(plot), group = tree)) +
  geom_point() +
  geom_line() +
  theme(legend.position = "none")
```

### Use midpoint

Assume we don't know the death time exactly

```{r}
#figure out in which inventory trees are marked dead
dead_year_records <- periodic |> 
  mutate(first_dead = time[which.max(is.na(ht))]) |> 
  mutate(alive = time < first_dead) |> 
  select(tree, plot, time, alive)


interpolated_midpt <-
  left_join(periodic_expanded, dead_year_records) |> 
  #linearly interpolate ht
  group_by(plot, tree) |> 
  mutate(ht = inter_extra_polate(time, ht)) |> 
  #use midpoint rule for death
  mutate(alive = step_interp(alive)) |> 
  filter(alive)
```

```{r}
ggplot(interpolated_midpt, aes(x = time, y = ht, color = as.factor(plot), group = tree)) +
  geom_point() +
  geom_line() +
  theme(legend.position = "none")
```

## Get population averages

I'm not yet entirely sure how to do this for the panel design, but for the simulated and interpolated data, you just take means.

```{r}
periodic_stat <- periodic |> 
  group_by(time) |> 
  summarize(
    # sum = sum(ht, na.rm = TRUE),
    mean = mean(ht, na.rm = TRUE),
    sd = sd(ht, na.rm = TRUE),
    sem = sd/sqrt(n())
  )
sim_stat <- sim |> 
  group_by(time) |> 
  summarize(
    sum = sum(ht, na.rm = TRUE),
    mean = mean(ht, na.rm = TRUE),
    sd = sd(ht, na.rm = TRUE),
    sem = sd/sqrt(n())
  )
interpolated_mortyr_stat <- interpolated_mortyr |> 
  group_by(time) |> 
  summarize(
    sum = sum(ht, na.rm = TRUE),
    mean = mean(ht, na.rm = TRUE),
    sd = sd(ht, na.rm = TRUE),
    sem = sd/sqrt(n())
  )
interpolated_midpt_stat <- interpolated_midpt |> 
  group_by(time) |> 
  summarize(
    sum = sum(ht, na.rm = TRUE),
    mean = mean(ht, na.rm = TRUE),
    sd = sd(ht, na.rm = TRUE),
    sem = sd/sqrt(n())
  )
```

```{r}
p <- position_dodge(width = 0.4)

list(
  sim = sim_stat,
  periodic = periodic_stat,
  interpolated_midpt = interpolated_midpt_stat,
  interpolated_mortyr = interpolated_mortyr_stat
) |> 
  bind_rows(.id = "method") |> 
  ggplot(aes(x = time, y = mean, color = method)) +
  geom_point(position = p) +
  geom_line(position = p) +
  geom_ribbon(aes(ymin = mean - sem, ymax = mean + sem, fill = method), 
              color = NA, alpha = 0.2, position = p) +
  # scale_x_continuous(limits = c(40, 55)) +
  NULL
```
