---
title: "Working with forestTIME-builder data"
author: Eric Scott
format: gfm
---

## Setup

Packages you'll need:

```{r}
#| message: false
library(dplyr)
library(dbplyr) #need to install, but don't actually need to explicitly load
library(DBI)
library(duckdb)
```

If any are missing, install with `install.packages("duckdb")`, for example.

## Step 1: connecting to the database

::: callout-tip
The path to the duckdb file can be a relative path if you're in an RStudio project.
If you're working in a Quarto or RMarkdown document, you might need to use `here::here("project/path/to/foresttime-DE.duckdb")` to ensure the same working directory whether you're working interactively or rendering the document.
:::

```{r}
#| echo: false
con <- dbConnect(duckdb(here::here("data/db/foresttime-DE.duckdb")))
```

```{r}
#| eval: false
con <- dbConnect(duckdb("data/db/foresttime-DE.duckdb"))
```

## Step 2: getting a table

List the tables in the database:

```{r}
dbListTables(con)
```

Pull a specific table to work with:

```{r}
tree_annualized <- tbl(con, "tree_annualized")
tree_annualized
```

## Step 3: working with tables

The table above is not in memory.
You can tell because it lists the dimensions as `[?? x 13]`—it doesn't know how many rows there are because it's only read in enough to figure out what kinds of data (numeric, character, date, etc.) the columns hold.
However, you can still use *most* tidyverse functions to manipulate it.

```{r}
tree_annualized |> 
  filter(DIA_est_midpoint < 5) |> 
  mutate(is_dead = YEAR >= midpoint_dead_year) |> 
  select(-STATECD)
```

This includes joins

```{r}
tree <- tbl(con, "tree")
plot <- tbl(con, "plot")
left_join(
  tree,
  plot |> select(PLOT_COMPOSITE_ID, DESIGNCD),
  by = join_by(PLOT_COMPOSITE_ID)
)
```

If you need to bring the dataframe into memory for some reason, you can use `collect()`.

```{r}
tiny_saplings <- tree |> 
  filter(DIA < 2) |> 
  collect()
tiny_saplings
```

The resulting dataframe is in memory and has 1226 rows.

::: callout-important
Be careful using `collect()` on very large tables.
You might crash your R session!
If you're just wanting to `View()` a table, you might consider using `head()` or `slice_sample()` on the data frame before using `collect()`
:::

```{r}
#| eval: false

tree |> 
  slice_sample(n = 100) |> #get 100 random rows
  collect() |> #pull into memory
  View() #open in data viewer tab in RStudio

```

## Step 4: disconnect

You can disconnect from the database with `dbDisconnect()`.
It isn't totally necessary and restarting the R session will do the same thing, but it's good practice to include in a script.

```{r}
dbDisconnect(con)
```
